---
title: "Flow Summary Analysis"
output: html_notebook
---
Author: Kevin Keating
Last updated: May 8 2023

# Read in data

Clear environment, add libraries
```{r, echo=FALSE}
rm(list = ls())
library(ggplot2)
library(ggrepel)
library(scales)
library(forcats)
library(munsell)
library(ggh4x)
library(wesanderson)
library(jtools)
library(gridExtra)
library(PerformanceAnalytics)
```

Provide info for all runs to be analyzed
```{r}
# Write in paths for all CSVs to be analyzed
CSVs = c("../run_data/e_coli_8.23.22_all_events.csv","../run_data/p_putida_all_events.csv","../run_data/c_necator_2.9.22_all_events.csv", "../run_data/c_necator_2.10.22_all_events.csv", "../run_data/c_necator_2.11.22_all_events.csv", "../run_data/k_nataicola_2.23.22_all_events.csv", "../run_data/k_nataicola_2.23.22_2_all_events.csv", "../run_data/k_nataicola_2.24.22_all_events.csv", "../run_data/k_nataicola_2.24.22_2_all_events.csv","../run_data/k_nataicola_2.24.22_3_all_events.csv")

# Provide filename to convert sample ID in output to meaningful ID
sample_names = read.csv("../run_data/sample_names.csv", header=TRUE)

# Remove DOE oversampling replicates - some conditions had extra replicates for DOE analysis. To keep the number of samples the same for all conditions, only the first set of replicates for each condition will be analyzed.
sample_names <- sample_names[!(sample_names$well == "H2" | sample_names$well == "B3" | sample_names$well == "D3" | sample_names$well == "E2"),]

# Create a list of the species for each filename in CSVs, then pair them in a dataframe
species_list = c("e_coli", "p_putida", "c_necator", "c_necator", "c_necator", "k_nataicola", "k_nataicola", "k_nataicola", "k_nataicola", "k_nataicola")
species_df <- data.frame(CSVs, species_list)

# Initialize dataframe to be filled in from parsed CSVs
df <- data.frame('promoter' = NA, 'rbs' = NA, 'replicate' = NA, 'run_id' = NA, 'species' = NA, 'MEFL' = NA, 'FSC' = NA, 'SSC' = NA)
```

Read in csv output of FlowCal analysis for all runs, parse and save to a dataframe

```{r}
for (file in CSVs) {
  temp_df = read.csv(file, header=TRUE)
  
  # Need to translate "sample" from FlowCal output into promoter + RBS + replicate columns
  # First add promoter and rbs columns from sample_names df
  for (row in 1:nrow(sample_names)) {
    temp_df$promoter[grepl(sample_names$well[row], temp_df$sample, fixed = TRUE)] <- sample_names$promoter[row]
    temp_df$rbs[grepl(sample_names$well[row], temp_df$sample, fixed = TRUE)] <- sample_names$rbs[row]
  }
  # Pull out replicate number from flag -rep
  nrep = 3
  for (rep in 1:nrep){
    temp_df$replicate[grepl(paste(c("-",rep),collapse=""), temp_df$sample, fixed = TRUE)] <- rep
  }
  # Add species
  temp_df$species <- species_df$species_list[grepl(file,species_df$CSVs, fixed = TRUE)]
  
  # Add parsed data to overall df and remove temp df
  df <- rbind(df,subset(temp_df, select = -sample))
  rm(temp_df)
}

rm(rep,nrep,row,file)
# Remove pesky initial NA row from df
df <- df[complete.cases(df),]
```

Add growth defect information to dataframe

```{r}
# Make dataframe of samples that have growth defects
defect_spec <- c(rep("c_necator",times=5),rep("k_nataicola",times=7))
defect_pro <- c("J23106","J23103","J23103","J23100","J23100","J23103","J23103","J23103","J23106","J23103","J23106","J23106")
defect_rbs <- c("RiboJ_BCD-2","RiboJ_BCD-2","B0033m","RiboJ_UTR1","RiboJ_BCD-2","B0033m","B0034m","BCD-2","B0033m","B0032m","B0034m","B0032m")
defects <- data.frame(defect_spec,defect_pro,defect_rbs)
rm(defect_spec,defect_pro,defect_rbs)


# Add this info to main dataframe
df$growth_defect <- "no"
for (i in 1:length(defects$defect_pro)) {
        df$growth_defect[df$species == defects$defect_spec[i] & df$promoter == defects$defect_pro[i] & df$rbs == defects$defect_rbs[i]] <- "yes"
}
rm(i)
```

Set factors as factors

```{r}
df$promoter <- as.factor(df$promoter)
df$rbs <- as.factor(df$rbs)
df$replicate <- as.factor(df$replicate)
df$run_id <- as.factor(df$run_id)
df$species <- as.factor(df$species)
```

Add column for MEFL with average empty vector MEFL subtracted, per run

```{r}
for (run in unique(df$run_id)){
  mean_autofluorescence = mean(subset(df, run_id==run & promoter=='empty_vector')$MEFL)
  df$MEFL_no_auto <- df$MEFL - mean_autofluorescence
  rm(mean_autofluorescence)
}
rm(run)
```

# Violin plots

Make violin plots for figure 2
```{r}
# Make species lookup list
spec.labs = c("e_coli" = "E. coli", "p_putida" = "P.putida", "c_necator" = "C. necator", "k_nataicola" = "K. nataicola")

# Define plot settings
for (specie in unique(df$species)){
  # Define starting data
  p <- ggplot(data=subset(df,replicate==1 & species == specie), aes(x=rbs, y=MEFL), axis.ticks.x=element_blank())
  
  # Format axes
  p <- p + xlab(" ") + ylab("Fluorescence, MEFL") + theme(axis.text.x = element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_text(size=20), axis.title.y=element_text(size=20,face="bold"), strip.text.x = element_text(size=20))
  p <- p + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x), labels = trans_format("log10", math_format(10^.x)), limits = c(10**-0.75,10**6.5))
  
  # + ggtitle(paste("Fluorescent intensity", spec.labs[specie]))
  
  # Assign color to replicate 1
  base_color <- "5B 5/6"
  p <- p + geom_violin(colour=mnsl(base_color), fill=mnsl(lighter(base_color, 2)), outlier.colour="transparent")
  
  # Add replicates 2 and 3
  p <- p + geom_violin(data = subset(df,replicate==2 & species == specie),colour = mnsl(rygbp(base_color, 2)),fill=mnsl(lighter(rygbp(base_color, 2),2)), outlier.colour="transparent")
  p <- p + geom_violin(data = subset(df,replicate==3 & species == specie),colour = mnsl(pbgyr(base_color, 2)),fill=mnsl(lighter(pbgyr(base_color, 2),2)), outlier.colour="transparent")

  # Facet labels
  pro.labs <- c("EV", "J23103", "J23106", "J23100")
  names(pro.labs) <- c("empty_vector","J23103","J23106","J23100") 
  
  # Make facets
  p <- p  +  facet_grid2(~fct_relevel(promoter,"empty_vector","J23103","J23106","J23100"),labeller = as_labeller(pro.labs), scales = "free", space = "free", strip = strip_vanilla(clip = "off",size = "variable"))

  # Display Plot
  #p
  # Save plot
  setwd("../outputs")
  ggsave(paste(c(specie,"violins.pdf"),collapse="_"), p, device = "pdf", width = 7, height = 5, units ="in")
}
```


# Lookup tables

Make summary / lookup table of all numbers, including negative values

```{r}
# Make new dataframe containing arithmetic mean of each promoter/rbs pair, pooling replicates
df_summary = setNames(data.frame(matrix(ncol = 6, nrow = 0)), c("mean","std_err", "species","rbs","promoter","growth_defect"))

for (specie in unique(df$species)) {
  for (pro in unique(df$promoter[! df$promoter %in% "empty_vector"])) {
      for (rb in unique(df$rbs[! df$rbs %in% "empty_vector"])) {
        df_summary_row <- cbind(data.frame("mean" = mean(subset(df,df$species==specie & df$promoter==pro & df$rbs ==rb)$MEFL)), data.frame("std_err" = sd(subset(df,df$species==specie & df$promoter==pro & df$rbs ==rb)$MEFL)/sqrt(length(subset(df,df$species==specie & df$promoter==pro & df$rbs ==rb)$MEFL))), data.frame("species" = specie), data.frame("rbs" = rb), data.frame("promoter" = pro), data.frame("growth_defect" = unique(df$growth_defect[df$species==specie & df$promoter==pro & df$rbs==rb])))
        df_summary = rbind(df_summary,df_summary_row)
      }
  }
}

rm(df_summary_row)

df_summary$species <- as.factor(df_summary$species)
df_summary$growth_defect <- as.factor(df_summary$growth_defect)

write.csv(df_summary,"../outputs/mean_summary.csv",row.names = FALSE)
```


Calculate 95% CI of means and add them to summary df
```{r}
alpha = 0.05
df_summary$CI_95_LB <- df_summary$mean - qnorm(1-alpha/2)*df_summary$std_err
df_summary$CI_95_UB <- df_summary$mean + qnorm(1-alpha/2)*df_summary$std_err
```

Make plots of mean - this generates figure 4 for openCIDAR paper
```{r}
# Put in data
p <- ggplot(df_summary, aes(x=fct_relevel(promoter, "J23103", "J23106", "J23100"), y=mean, label = rbs, fill = growth_defect)) 

# Labeling stuff
spec.labs <- c("E. coli", "P. putida", "C. necator", "K. nataicola")
names(spec.labs) <- c("e_coli", "p_putida", "c_necator", "k_nataicola")

# Add formatting
p <- p + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x))) + geom_point(size=2, shape=23) + facet_wrap(~fct_relevel(species,"e_coli","p_putida","c_necator","k_nataicola"), ncol=4, labeller = as_labeller(spec.labs)) + geom_text_repel(size=2) + xlab("Promoter") + ylab("Mean MEFL") + guides(fill = guide_legend(title = "Growth defect", reverse = TRUE)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Save and display plot
ggsave(filename="../outputs/arithmetic_mean_LUT.pdf")
p
```

Make separate dataframe for positive values of MEFL only (negative values result from conversion of voltage -> AU -> MEFL)

```{r}
# Pull out a new df of only positive values
pos_df <- subset(df, df$MEFL > 0)
```


Make summary / lookup table of positive numbers only

```{r}
# Make new dataframe containing geometric mean of each promoter/rbs pair, pooling replicates
pos_df_summary = setNames(data.frame(matrix(ncol = 7, nrow = 0)), c("mean","std_err","gmean", "species","rbs","promoter","growth_defect"))

for (specie in unique(pos_df$species)) {
  for (pro in unique(pos_df$promoter[! pos_df$promoter %in% "empty_vector"])) {
      for (rb in unique(pos_df$rbs[! pos_df$rbs %in% "empty_vector"])) {
        pos_df_summary_row <- cbind(data.frame("mean" = mean(subset(pos_df,pos_df$species==specie & pos_df$promoter==pro & pos_df$rbs ==rb)$MEFL)), data.frame("std_err" = sd(subset(pos_df,pos_df$species==specie & pos_df$promoter==pro & pos_df$rbs ==rb)$MEFL)/sqrt(length(subset(pos_df,pos_df$species==specie & pos_df$promoter==pro & pos_df$rbs ==rb)$MEFL))), data.frame("gmean" = exp(mean(log(subset(pos_df,pos_df$species==specie & pos_df$promoter==pro & pos_df$rbs ==rb)$MEFL)))), data.frame("species" = specie), data.frame("rbs" = rb), data.frame("promoter" = pro), data.frame("growth_defect" = unique(pos_df$growth_defect[pos_df$species==specie & pos_df$promoter==pro & pos_df$rbs==rb])))
        pos_df_summary = rbind(pos_df_summary,pos_df_summary_row)
      }
  }
}

rm(pos_df_summary_row)

pos_df_summary$species <- as.factor(pos_df_summary$species)
pos_df_summary$growth_defect <- as.factor(pos_df_summary$growth_defect)

write.csv(pos_df_summary,"../outputs/gmean_summary.csv",row.names = FALSE)
```


# Original CIDAR comparison

Import original CIDAR study values and graph against openCIDAR constructs

```{r}
# Import original CIDAR values
og_cidar = read.csv("../run_data/CIDAR_MEFL_values.csv", header=TRUE)
```

Make plot (for Figure 2B)

```{r}
# Put in data
p2 <- ggplot(og_cidar, aes(x=rbs, y=gmean, label = rbs)) 

# Add formatting
base_color <- "5B 5/6"

p2 <- p2 + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x), labels = trans_format("log10", math_format(10^.x)), limits = c(10**-0.75,10**6.5)) + geom_point(size=6, shape=23, colour=mnsl(base_color), fill=mnsl(lighter(base_color, 2)), outlier.colour="transparent") + xlab(" ") + ylab("Fluorescence, MEFL") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + facet_wrap(~fct_relevel(promoter,"J23103","J23106","J23100"), ncol=3) + theme(axis.text.x = element_blank(), axis.ticks.x=element_blank(), axis.text.y=element_text(size=20), axis.title.y=element_text(size=20,face="bold"), strip.text.x = element_text(size=20))


# Save and display plot
ggsave(filename="../outputs/og_CIDAR_comparison.pdf")
p2
```

# Rank order plots

Make rank roder plot for figure 3A

```{r}
# Add design column to summary row to combine promoter / RBS combination together into one variable
pos_df_summary$design <- paste(pos_df_summary$promoter,pos_df_summary$rbs,sep="_")
pos_df_summary$design <- as.factor(pos_df_summary$design)

# Make list 
pos_ec_sum = subset(pos_df_summary,pos_df_summary$species=="e_coli")
pos_ec_rank_order = reorder(pos_ec_sum$design,pos_ec_sum$gmean)

# Put in data
p <- ggplot(pos_df_summary, aes(x=fct_relevel(design,levels(pos_ec_rank_order)), y=gmean, fill = species))
p <- p + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x), labels = trans_format("log10", math_format(10^.x))) + geom_point(size=2, shape=23) + ggtitle("Rank order comparison") + xlab("Promoter / RBS pair") + ylab("Arithmetic mean MEFL") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + scale_fill_discrete(name = "Species", labels = c("C. necator", "E. coli", "K. nataicola", "P. putida"))

ggsave(filename="../outputs/gmean_rank_order_comparison.pdf")
p
```

# Correlation plots
Log-transform data

```{r}
log_gmean_corr_tb <- data.frame(E.coli = log10(pos_df_summary$gmean[pos_df_summary$species=="e_coli"]), P.putida = log10(pos_df_summary$gmean[pos_df_summary$species=="p_putida"]), C.necator = log10(pos_df_summary$gmean[pos_df_summary$species=="c_necator"]), K.nataicola = log10(pos_df_summary$gmean[pos_df_summary$species=="k_nataicola"]))

chart.Correlation(log_gmean_corr_tb, histogram=FALSE, pch=19)
```

Correlation plot for figure 3B

```{r}
# Make correlation panel figure
panel.cor <- function(x, y){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- round(cor(x, y), digits=2)
    txt <- paste0("R = ", r)
    text(0.5, 0.5, txt, cex = 2)
}
# Customize upper panel
upper.panel<-function(x, y){
  points(x,y, pch = 19)
}
# Create the plots
pdf("../outputs/correlation_plot.pdf")
pairs(log_gmean_corr_tb, 
      lower.panel = panel.cor,
      upper.panel = upper.panel)
dev.off()
```

What happens when you include only J23103 and J23106 in the dataset? This script produces supp. fig 5

```{r}
log_gmean_corr_tb_no_J23100 <- data.frame(E.coli = log10(pos_df_summary$gmean[pos_df_summary$species=="e_coli" & pos_df_summary$promoter != "J23100"]), P.putida = log10(pos_df_summary$gmean[pos_df_summary$species=="p_putida" & pos_df_summary$promoter != "J23100"]), C.necator = log10(pos_df_summary$gmean[pos_df_summary$species=="c_necator" & pos_df_summary$promoter != "J23100"]), K.nataicola = log10(pos_df_summary$gmean[pos_df_summary$species=="k_nataicola" & pos_df_summary$promoter != "J23100"]))

chart.Correlation(log_gmean_corr_tb_no_J23100, histogram=FALSE, pch=19)

# Make correlation panel figure for paper
panel.cor <- function(x, y){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- round(cor(x, y), digits=2)
    txt <- paste0("R = ", r)
    text(0.5, 0.5, txt, cex = 2)
}
# Customize upper panel
upper.panel<-function(x, y){
  points(x,y, pch = 19)
}
# Create the plots
pdf("../outputs/correlation_plot_no_J23100.pdf")
pairs(log_gmean_corr_tb_no_J23100, 
      lower.panel = panel.cor,
      upper.panel = upper.panel)
dev.off()
```

# Linear models

Build linear models to describe data

Note - here are some useful functions

print()	simple printed display
summary()	standard regression output
coef()	(or coefficients()) extracting the regression coefficients
residuals()	(or resid()) extracting residuals
fitted()	(or fitted.values()) extracting fitted values
anova()	comparison of nested models
predict()	predictions for new data
plot()	diagnostic plots
confint()	confidence intervals for the regression coefficients
deviance()	residual sum of squares
vcov()	(estimated) variance-covariance matrix
logLik()	log-likelihood (assuming normally distribted errors)
AIC	information criteria including AIC, BIC/SBC (assuming normally distributed errors)


```{r}
# Make sure "empty_vector" is first level for promoter and rbs
pos_df$rbs <- relevel(pos_df$rbs,"empty_vector")
pos_df$promoter <- relevel(pos_df$promoter,"empty_vector")
```

Construct linear models with interactions - these are the ones used in figure 5

```{r}
ec_pro_rbs_int_lm <-lm(log10(MEFL) ~ promoter * rbs, data = subset(pos_df, pos_df$species == "e_coli"))
summary(ec_pro_rbs_int_lm)

pp_pro_rbs_int_lm <-lm(log10(MEFL) ~ promoter * rbs, data = subset(pos_df, pos_df$species == "p_putida"))
summary(pp_pro_rbs_int_lm)

cn_pro_rbs_int_lm <-lm(log10(MEFL) ~ promoter * rbs, data = subset(pos_df, pos_df$species == "c_necator"))
summary(cn_pro_rbs_int_lm)

kn_pro_rbs_int_lm <-lm(log10(MEFL) ~ promoter * rbs, data = subset(pos_df, pos_df$species == "k_nataicola"))
summary(kn_pro_rbs_int_lm)
```

Data wrangling
```{r}
types <- c("intercept", rep("promoter",each=3), rep("RBS",each=5), rep("interaction",each=10))

ec_est <- data.frame(summary(ec_pro_rbs_int_lm)$coefficients)
ec_est$parameter <- row.names(ec_est)
ec_est$org <- "e_coli"
ec_est <- cbind.data.frame(ec_est,types)
pp_est <- data.frame(summary(pp_pro_rbs_int_lm)$coefficients)
pp_est$parameter <- row.names(pp_est)
pp_est$org <- "p_putida"
pp_est <- cbind.data.frame(pp_est,types)
cn_est <- data.frame(summary(cn_pro_rbs_int_lm)$coefficients)
cn_est$parameter <- row.names(cn_est)
cn_est$org <- "c_necator"
cn_est <- cbind.data.frame(cn_est,types)
kn_est <- data.frame(summary(kn_pro_rbs_int_lm)$coefficients)
kn_est$parameter <- row.names(kn_est)
kn_est$org <- "k_nataicola"
kn_est <- cbind.data.frame(kn_est,types)

lm_est <- rbind.data.frame(ec_est,pp_est,cn_est,kn_est)
```

Plots - figure 5
```{r}
sspec.labs <- c("E. coli", "P. putida", "C. necator", "K. nataicola")
names(spec.labs) <- c("e_coli", "p_putida", "c_necator", "k_nataicola")

pal <- wes_palette("Darjeeling1",4,"discrete")

p1 <- ggplot(subset(lm_est,types=="intercept"), aes(x=parameter, y=Estimate)) + geom_bar(stat="identity", fill=pal[1]) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none") + facet_wrap(~fct_relevel(org,"e_coli","p_putida","c_necator","k_nataicola"), ncol=4, labeller = as_labeller(spec.labs))

p2 <- ggplot(subset(lm_est,types=="promoter"), aes(x=parameter, y=Estimate)) + geom_bar(stat="identity", fill=pal[2]) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none") + facet_wrap(~fct_relevel(org,"e_coli","p_putida","c_necator","k_nataicola"), ncol=4, labeller = as_labeller(spec.labs))

p3 <- ggplot(subset(lm_est,types=="RBS"), aes(x=parameter, y=Estimate)) + geom_bar(stat="identity", fill=pal[3]) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none") + facet_wrap(~fct_relevel(org,"e_coli","p_putida","c_necator","k_nataicola"), ncol=4, labeller = as_labeller(spec.labs))

p4 <- ggplot(subset(lm_est,types=="interaction"), aes(x=parameter, y=Estimate)) + geom_bar(stat="identity", fill=pal[4]) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position = "none") + facet_wrap(~fct_relevel(org,"e_coli","p_putida","c_necator","k_nataicola"), ncol=4, labeller = as_labeller(spec.labs))

pdf("../outputs/LM_4x4.pdf")
grid.arrange(p1, p2, p3, p4, nrow=2)
dev.off()
```

